Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-12 paper/paper.tex
scattered edits

%notation
We use the following notation: $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise
complex conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate.

@@

A random vector $z\in \cplx^m$ has complex Gaussian distribution, $
  z \sim q(z) = \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
$, with mean $\mu \in \cplx^m$ and $\cplx^{m\times m}$ covariance and relation matrices

@@

@@

@@

provided they satisfy $\Gamma \succeq 0$, $\Gamma^{\hop} = \Gamma$, $C^\top = C$, and $

@@

$C = 0$. The entropy of a proper $\cplx$-Gaussian is

@@

The local reparameterization trick provides both faster computations, and better statistical
efficiency of the SGVB gradient estimator, \citep{wang_fast_2013,kingma_variational_2015}.
We assume that weights $W \in \cplx^{n\times m}$ of a dense linear layer are drawn from a
mean field $\cplx$-Gaussian approximate posterior
\begin{equation}  \label{eq:c-gauss-vi-general}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij}, \Sigma_{ij}, C_{ij}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and \textit{relation} are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \in \real$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. Property \eqref{eq:cn-affine} and independence of $W$ imply that for any $x \in \cplx^m$
and $b \in \cplx^n$ the output $y = W x + b$ is distributed like
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
The trick requires three matrix operations: $\mu x + b$, $\Sigma \lvert x \rvert^2$ and
$C x^2$ with the modulus and power applied elementwise. For a proper approximation
\eqref{eq:c-gauss-vi} the relation coefficients are zero.

The trick and dropout can be applied to any layer, the output of which depends linearly
on its parameters, e.g. convolutional layers, affine and bilinear transformations ($
  y_j = x^\top W^{(j)} z + b_j
$). Similar to $\real$ case, variational dropout for $\cplx$ convolutions draws independent
kernel $W$ for each spatial patch in the input, since a convolution is a matrix product
of $W$, unrolled into a toeplitz matrix, and flattened intput $\vec{x}$. Weight sharing
is essentially pushed to shared variational parameters $\theta$ in the approximation $q_\theta$.
The rationale is to reduce correlation in the gradient estimator stemming from kernel weights
being shared between overlapping patches, \citep{kingma_variational_2015,citation_needed}.
Thus \eqref{eq:cplx-gauss-trick} requires computing $\cplx$ convolutions of $x$, complex
square $x^2$, and amplitude $\lvert x \rvert^2$ with the mean, relation, and variance kernels,
respectively.
%
\todo{other reasons for patch-independent weights}

For $\cplx$-variational dropout we propose using mean-field proper $\cplx$-Gaussian approximate
posterior, e.g. parameters $W \in \cplx^{n\times m}$ of a dense linear layer follow
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W \sim q(W)
    % = \prod_{ij} q(W_{ij})
    = \prod_{ij} \mathcal{N}^{\cplx}(W_{ij} \mid \mu_{ij}, \sigma^2_{ij}, 0)
  \,,
\end{equation}
and we use similar factorized approximation for convolutions. We propose to MAP estimates,
of the biases i.e. assume an independent fully factorized Dirac approximation.

@@

@@

Consider the following parameterization of $W$: $
  W = \mu \odot \xi
$ where $
  \xi_{ij} \sim \mathcal{N}(1, \alpha_{ij})
$, yet $\mu \in \cplx^{n \times m}$. This case corresponds to inference regarding
relevance coefficients $\xi$ rather than the parameters themselves after. Under this
parameterization
\begin{equation}  \label{eq:real-scal-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij},
      \, \alpha_{ij} \lvert \mu_{ij} \rvert^2,
      \, \alpha_{ij} \mu_{ij}^2
    \bigr)
  \,,
\end{equation}
which is a degenerate $\cplx$-Gaussian distribution. \eqref{eq:cplx-gauss-trick} implies
\begin{equation}  \label{eq:real-scal-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}\bigl(
      e_i^\top \mu x + b,
      \, \sum_j \alpha_{ij} \lvert x_{ij} \mu_{ij}\rvert^2,
      \, \sum_j \alpha_{ij} (x_{ij} \mu_{ij})^2,
      \,
    \bigr)
    \,.
\end{equation}
The KL-divergence term in \eqref{eq:elbo} coincides with \eqref{eq:improper-kl-div-real}.
We use the approximation provided by \citep{molchanov_variational_2017}. Note that additive
noise reparameterization, \citep{molchanov_variational_2017}, cannot be used in
\eqref{eq:real-scal-gauss-trick}.

@@

%notation
By $e_i$ we denote the $i$-th real unit vector of dimensionality \textit{conforming} to
the matrix-vector expression it is used in. $\vec{M}$ denotes \textit{row-major} flattening
of $M$ into a vector, i.e. in lexicographic order of its indices. $\diag{(\cdot)}$ embeds
$\cplx^n$ into $n\times n$ diagonal matrices. $\otimes$ denotes the Kronecker product,
for which we note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$, \citep{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.

If we assume a mean field $\cplx$-Gaussian approximate posteriror for $
  W \in \cplx^{n\times m}
$, then
\begin{equation}  \label{eq:c-gauss-vi-general}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagonal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
Hence \eqref{eq:cplx-gauss-trick} folows.

>>>>

@@

%notation
In the following $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise complex
conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate.

A vector $z\in \cplx^m$ has complex Gaussian distribution, $
  q(z) = \mathcal{N}^{\cplx}_m(\mu, \Gamma, C)
$ with mean $\mu \in \cplx^m$ and $\cplx^{m\times m}$ covariance and relation matrices

@@

  \begin{pmatrix}

@@

  \end{pmatrix}

@@

provided $\Gamma \succeq 0$, $\Gamma^{\hop} = \Gamma$, $C^\top = C$, and $

@@

$C = 0$. From \eqref{eq:c-gauss-entropy-derivation} its entropy is

@@

The local reparameterization trick provides both faster computations, and better statistical
efficiency of the SGVB gradient estimator, \citep{wang_fast_2013,kingma_variational_2015}.
We assume that weights $W \in \cplx^{n\times m}$ of a dense linear layer are drawn from a
mean field $\cplx$-Gaussian approximate posterior
\begin{equation}  \label{eq:c-gauss-vi-general}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij}, \Sigma_{ij}, C_{ij}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and \textit{relation} are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \in \real$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. Property \eqref{eq:cn-affine} and independence of $W$ imply that for any $x \in \cplx^m$
and $b \in \cplx^n$ the components of $y = W x + b$ are independent with
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \Bigl(
        % e_i^\top \mu x + b_i,
        b_i + \sum_{j=1}^m \mu_{ij} x_j,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \Bigr)
    \,.
\end{equation}
The trick requires three matrix operations: $\mu x + b$, $\Sigma \lvert x \rvert^2$ and
$C x^2$ with the modulus and power applied elementwise.

The trick and dropout can be applied to any layer, the output of which depends linearly
on its parameters, e.g. convolutional layers, affine and bilinear transformations $
  (x, z) \mapsto x^\top W^{(j)} z + b_j
$. Similar to $\real$ case, variational dropout for $\cplx$ convolutions draws independent
kernel $W$ for each spatial patch in the input, since a convolution is a matrix product
of $W$, unrolled into a toeplitz matrix, and flattened intput $x$. Weight sharing is
essentially pushed to shared variational parameters $\theta$ in the approximation $q_\theta$.
The rationale is to reduce correlation in the gradient estimator stemming from kernel weights
being shared between overlapping patches, \citep{kingma_variational_2015,citation_needed}.
Thus \eqref{eq:cplx-gauss-trick} requires computing $\cplx$ convolutions of $x$, complex
square $x^2$, and amplitude $\lvert x \rvert^2$ with the mean, relation, and variance kernels,
respectively.
%
\todo{other reasons for patch-independent weights}

@@

For $\cplx$-variational dropout we propose using mean-field proper $\cplx$-Gaussian approximate
posterior, e.g. parameters $W \in \cplx^{n\times m}$ of a dense linear layer follow
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W \sim q(W)
    % = \prod_{ij} q(W_{ij})
    = \prod_{ij} \mathcal{N}^{\cplx}(
      W_{ij} \mid
        \mu_{ij},
        \alpha_{ij} \lvert \mu_{ij} \rvert^2,
        % \sigma^2_{ij},
        0
    )
  \,,
\end{equation}
and we use similar factorized approximation for convolutions. For this approximation the
relation coefficients \eqref{eq:cplx-gauss-trick} are zero. We use MAP estimates of bias
terms, i.e. assume an independent fully factorized Dirac approximation.

@@

Consider the following parameterization of $W$: $
  W = \mu \odot \xi
$ where $
  \xi_{ij} \sim \mathcal{N}(1, \alpha_{ij})
$, yet $\mu \in \cplx^{n \times m}$. This case corresponds to inference regarding
relevance coefficients $\xi$ rather than the parameters themselves after. Under this
parameterization
\begin{equation}  \label{eq:real-scal-gauss-vi}
  W_{ij}
    \sim \mathcal{N}^{\cplx} \bigl(
      \mu_{ij},
      \, \alpha_{ij} \lvert \mu_{ij} \rvert^2,
      \, \alpha_{ij} \mu_{ij}^2
    \bigr)
  \,,
\end{equation}
which is a degenerate $\cplx$-Gaussian distribution. \eqref{eq:cplx-gauss-trick} implies
\begin{equation}  \label{eq:real-scal-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}\bigl(
      % e_i^\top \mu x + b,
      b_i + \sum_{j=1}^m \mu_{ij} x_j,
      \, \sum_j \alpha_{ij} \lvert x_{ij} \mu_{ij}\rvert^2,
      \, \sum_j \alpha_{ij} (x_{ij} \mu_{ij})^2
    \bigr)
    \,.
\end{equation}
The KL-divergence term in \eqref{eq:elbo} coincides with \eqref{eq:improper-kl-div-real}.
We use the approximation provided by \citep{molchanov_variational_2017}. Note that additive
noise reparameterization, \citep{molchanov_variational_2017}, cannot be used in
\eqref{eq:real-scal-gauss-trick}.

@@

%notation
By $e_i$ we denote the $i$-th real unit vector of dimensionality \textit{conforming} to the
matrix-vector expression it is used in. Let $\vec{M}$ denote \textit{row-major} flattening
of $M$ into a vector, i.e. in lexicographic order of its indices. $\diag{(\cdot)}$ embeds
$\cplx^n$ into $n\times n$ diagonal matrices, and $\otimes$ denotes the Kronecker product,
for which we note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$, \citep{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.

If we assume a mean field $\cplx$-Gaussian approximate posteriror for $
  W \in \cplx^{n\times m}
$, then
\begin{equation}  \label{eq:c-gauss-vi-general}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagonal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b_i,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
Hence \eqref{eq:cplx-gauss-trick} folows.
